Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-07 agda.tex
more progress

% !TEX root = hazelnut-popl17.tex
So far, we have given an overview of the most important judgements and rules in
the semantics of Hazelnut, and stated the important metatheorems. In a few
cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
have mechanized the semantics and metatheory of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
refer readers unfamiliar with Agda to the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.

This development is available in the supplemental material. At the time of submission, all of the rules as well as the proofs of Theorem \ref{thrm:actsafe} and Theorem \ref{thrm:actdet} (including the proofs a number of technical lemmas that we elided here) are complete.\todo{rest of it}

>>>>

% !TEX root = hazelnut-popl17.tex
So far, we have given an overview of the most important judgements and rules in
the semantics of Hazelnut, and stated the critical metatheorem, Sensibility, and several auxiliary ``checksum'' metatheorems. In a few
cases, we have informally sketched out why these metatheorems will hold. In order to formally establish that our design meets our stated objectives, we
have mechanized the semantics of Hazelnut using the Agda proof assistant \cite{norell:thesis}. We 
refer readers unfamiliar with Agda to the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.

This development is available in the supplemental material. At the time of submission, the full set of rules is available and the proofs of Theorem \ref{thrm:actsafe}, Theorem \ref{lemma:movement-erasure} and a number of technical lemmas that we elided or mentioned informally in the text are complete. We also have nearly all of the cases of Reachability and Constructability complete (and see no major barrier to the remaining cases -- they are simply tedious.) \todo{include link to repos in non-anonymous supplementals}\todo{talk about determinism either here or there}

### 2016-07-07 formally.tex
more progress

\subsubsection{Type Inconsistency}
In order to define the action semantics, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either direrctly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.

@@

The bidirectional action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}%\\\\
    % \zexp, \alpha~\mathsf{subsumes}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.  The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}).

@@

\begin{lemma}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px] 
\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
\end{enumerate}
\end{lemma}
\noindent Lemma \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.

@@

The simplest way to prove Theorem \ref{thrm:reachability} is to  break it into two lemmas. Lemma \ref{lemma:reach-up} establishes that you can always move the cursor to the outermost position in an expression. This serves as a check on our $\aMove{\dParent}$ rules.
\begin{lemma}[Reach Up]\label{lemma:reach-up} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$. 
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
\end{enumerate}
\end{lemma}
Lemma \ref{lemma:reach-down} establishes that you can always move the cursor from the outermost position to any other position. This serves as a check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules. 
\begin{lemma}[Reach Down]\label{lemma:reach-down} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
\end{enumerate}
\end{lemma}
Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. (The witness to the existential is not, of course, the shortest path between any two cursor locations.)

@@

\paragraph{Ascription} Moving on to expression actions, we start to see some more interesting rules. The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type. In the first case, the synthesized type appears in the ascription:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the type provided for analysis appears in the ascription:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}
% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the focused empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

@@

\begin{subequations}
For example, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
\begin{equation}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\htau}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

@@

When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
\begin{equation}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}

>>>>

\subsubsection{Type Inconsistency}
In some of the rules below, we will need to supplement our definition of type consistency from Figure \ref{fig:type-consistency} with a definition of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can define this notion either directly as the metatheoretic negation of type consistency, or as a separate inductively defined judgement. The supplement does the latter. The key rule establishes that arrow types are inconsistent with $\tnum$:
% \begin{subequations}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau}{\htau'}
  %   }{
  %     \tincompat{\htau'}{\htau}
  %   }
  % \end{equation}
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_1}{\htau_1'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
  % \begin{equation}
  %   \inferrule{
  %     \tincompat{\htau_2}{\htau_2'}
  %   }{
  %     \tincompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
  %   }
  % \end{equation}
% \end{subequations}
% The remaining rules, given in the appendix, establish that type incompatibility is symmetric.
The mechanization proves that this judgemental definition of type inconsistency is indeed the negation of type consistency.

@@

The action semantics includes a subsumption rule similar to the subsumption rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}%\\\\
    % \zexp, \alpha~\mathsf{subsumes}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type, then we defer to the synthetic action judgement, as long as the type of the resulting cursor erasure is consistent with the type provided for analysis.  The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem \ref{thrm:actsafe} goes through by induction and static subsumption, i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be the rule of last resort.

@@

\begin{theorem}[Movement Erasure Invariance]\label{lemma:movement-erasure} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px] 
\item If $\performMove{\ztau}{\aMove{\delta}}{\ztau'}$ then $\removeSel{\ztau}=\removeSel{\ztau'}$.
\item If $\performMove{\zexp}{\aMove{\delta}}{\zexp'}$ then $\removeSel{\zexp}=\removeSel{\zexp'}$.
\end{enumerate}
\end{theorem}
\noindent Theorem \ref{lemma:movement-erasure} is useful also in that the relevant cases of Theorem \ref{thrm:actsafe} are straightforward by its application.

@@

The simplest way to prove Theorem \ref{thrm:reachability} is to  break it into two lemmas. Lemma \ref{lemma:reach-up} establishes that you can always move the cursor to the outermost position in an expression. This serves as a check on our $\aMove{\dParent}$ rules.
\begin{lemma}[Reach Up]\label{lemma:reach-up} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\ztau}{\bar\alpha}{\zwsel{\htau}}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}$. 
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performSyn{\hGamma}{\zexp}{\htau}{\bar\alpha}{\zwsel{\hexp}}{\htau}$.
\end{enumerate}
\end{lemma}
Lemma \ref{lemma:reach-down} establishes that you can always move the cursor from the outermost position to any other position. This serves as a check on our $\aMove{\dChild}$ and $\aMove{\dNext}$ rules. 
\begin{lemma}[Reach Down]\label{lemma:reach-down} ~
\begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
\item If $\removeSel{\ztau}=\htau$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performTyp{\zwsel{\htau}}{\bar\alpha}{\ztau}$.
\item If $\hana{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ and $\performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}$.
\item If $\hsyn{\hGamma}{\hexp}{\htau}$ and $\removeSel{\zexp}=\hexp$ then there exists some $\bar\alpha$ such that $\bar\alpha~\mathsf{movements}$ nand $\performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\bar\alpha}{\zexp}{\htau}$.
\end{enumerate}
\end{lemma}
Theorem \ref{thrm:reachability} follows by straightforward composition of these two lemmas. %(The witness to the existential is not, of course, the shortest path between any two cursor locations.)

@@

\paragraph{Ascription} The $\aConstruct{\fasc}$ action operates differently depending on whether the H-expression under the cursor synthesizes a type or is being analyzed against a type. In the first case, the synthesized type appears in the ascription:
\begin{equation}
  \label{r:constructasc}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}{\htau}
  }
\end{equation}
In the second case, the type provided for analysis appears in the ascription:
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aConstruct{\fasc}}{\hexp : \zwsel{\htau}}
  }
\end{equation}
% This rule should supercede the subsumption rule, so we excluded $\aConstruct{\fasc}$ from subsumption in Rule (\ref{rule:subsumes}).

@@

\paragraph{Lambdas} The $\aConstruct{\flam{x}}$ action places a lambda abstraction binding $x$ into an empty hole. If the empty hole is being asked to synthesize a type, then the result of the action is a lambda ascribed the type $\tarr{\tehole}{\tehole}$, with the cursor on the argument type hole:
\begin{equation}
  \label{r:conelamhole}
  \inferrule{ }{
    \performSyn
      {\hGamma}
      {\zwsel{\hehole}}
      {\tehole}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\hehole} : \tarr{\zwsel{\tehole}}{\tehole}}
      {\tarr{\tehole}{\tehole}}
  }
\end{equation}
The type ascription is necessary because lambda expressions do not synthesize a type. If the empty hole is being analyzed against a type with matched arrow type, then no ascription is necessary:\todo{supp}
\begin{equation}\label{rule:performAna-lam-1}
  \inferrule{
    \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}
  }{
    \performAna
      {\hGamma}
      {\zwsel{\hehole}}
      {\htau}
      {\aConstruct{\flam{x}}}
      {\hlam{x}{\zwsel{\hehole}}}
  }
\end{equation}
% Because the ascription is omitted here but included in Rule (\ref{r:conelamhole}), we must exclude all $\aConstruct{\flam{x}}$ actions from subsumption in Rule (\ref{rule:subsumes}).

@@

\begin{subequations}
In particular, when the cursor is in the body of a lambda expression, the zipper case mirrors Rule (\ref{rule:syn-lam}):\todo{supp}
\begin{equation}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \performAna
    {\hGamma, x : \htau_1}
    {\zexp}
    {\htau_2}
    {\alpha}
    {\zexp'}
}{
  \performAna
    {\hGamma}
    {\hlam{x}{\zexp}}
    {\htau}
    {\alpha}
    {\hlam{x}{\zexp'}}
}
\end{equation}

@@

When the focus is in the type position of an ascription, we must re-check the ascribed expression because the cursor erasure might have changed (in practice, one would optimize this check to only occur if the cursor erasure did change):
\begin{equation}\label{rule:zipper-asc}
\inferrule{
  \performTyp{\ztau}{\alpha}{\ztau'}\\
  \hana{\hGamma}{\hexp}{\removeSel{\ztau'}}
}{
  \performSyn
    {\hGamma}
    {\hexp : \ztau}
    {\removeSel{\ztau}}
    {\alpha}
    {\hexp : \ztau'}
    {\removeSel{\ztau'}}
}
\end{equation}

### 2016-07-07 implementation.tex
added new impl figure; more closely matches terminology and syntax from paper
minor figure tweak: user moves in time
minor: bigger figure
tweaked title linebreak spacing; updated colors and stroke for hazel
final sec 4
more progress

\subsection{Implementation Concepts}

\begin{figure}
\centering
\includegraphics[width=4in]{Implementation_Diagram}
\caption{Implementation Concepts. Each Action--Zipper Structure--View combination is considered to appear ``instantaneously'' on the timeline.}
\label{fig:FRP}
\end{figure}

@@

A substantially simpler system that we developed while exploring the ideas that led to Hazelnut can be found at the following URL:
\url{http://www.cs.cmu.edu/~comar/nestedpairs/}.

>>>>

% !TEX root = hazelnut-popl17.tex
\subsection{Implementation Concepts}
Central to any implementation of Hazelnut is a stream of edit states whose cursor erasures synthesize types according to the synthetic action judgement under an empty context, $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$. The middle row of Figure \ref{fig:impl-overview} diagrams this stream of edit states. By Theorem \ref{thrm:actsafe}, the editor does not need to typecheck each edit state (though in some of the rules, e.g. Rule (\ref{rule:zipper-asc}) which handles the situation where the type in a type ascription changes, portions of the program do need to be typechecked.) For example, the reader is encouraged to re-examine the examples in Figure \ref{fig:first-example} and \ref{fig:second-example} -- the cursor erasure of each edit state synthesizes a type.

@@

\begin{figure}
\centering
\includegraphics[width=0.90\columnwidth]{impl-overview2}
\caption{...}
\label{fig:impl-overview}
\end{figure}
